\textbf{Automated Reasoning\\Practical Assignment -- Part 2}

### 2022-12-15 part2/report/paper.tex
Report on mosaic

@@

% section robot (end)
\appendix
\pagebreak
\section{Groups}
\label{apx: groups}
\begin{enumerate}[label=\alph*)]
	\item
		Smallest finite group for which $inv(x*y)=inv(x)*inv(y)$ does not hold.
		\begin{verbatim}
		interpretation( 6, [number=1, seconds=0], [

>>>>

\newcommand{\black}{\cellcolor{black}\color{white}}

@@

% section robot (end)

\pagebreak
\section{Mosaic} % (fold)
\label{sec:mosaic}
	\subsection{Puzzle} % (fold)
	\label{sub:puzzle}
	Our problem is to generate a minimal mosaic puzzle.
	Mosaic\footnote{Also known as: fill-a-pix} is a relatively simple puzzle.
	You are given a grid of squares in which some cells have a number.
	You should color each cell either white or black.
	A number indicates how many cells in the surrounding $3\times3$ should be colored black.
	Below you can find an example of a mosaic puzzle and its solution.

	\begin{figure}[H]
		\centering
		\includesvg[width=0.8\textwidth]{resources/mosaic}
	\end{figure}

	The puzzle we generate should satisfy the following requirements:
	\begin{enumerate}
		\item The puzzle has an unique solution.
		\item This solution should match the solution we provide as input.
		\item There are no other puzzles with the same solution that have fewer numbers given.
	\end{enumerate}

	We have made two programs to generate puzzles.
	The first uses a SMT solver iteratively to find constraints that are necessary for a puzzle to be unique. Then it runs another SMT solver to minimize the number of hints. The other approach uses the unsat core. While Z3 finds a small puzzle using this method, it is not guaranteed to satisfy (3).
	% subsection puzzle (end)

	\subsection{Generation using unsat core} % (fold)
	\label{sub:mosaic_unsat_core}
	Since we want to use the unsat core, we need to make a set of constraints that is not satisfiable. We will use this unsatisfiability to guarantee requirement (1), namely we will try to find a second solution.

	For this we add a variable for each cell, whether it is black in the second solution or not. Since we will be adding these later, we encode this as a 0-1 constrained Z3 integer:

	$$\forall_r,c:  0 \le \text{filled}_{r,c} \le 1$$

	Then we will add the constraint that at least one of those variable does not match the given solution (grid), so that we force Z3 to look for a second solution:

	$$\exists_{r,c}:  \text{filled}_{r,c} \ne \text{grid}_{r,c}$$

	We will ask Z3 to find this second solution to the puzzle with a number in every cell. Since we have a desired solution, we can determine which value the numbers should have. For every cell we add the constraint that it has the same number of black cells surrounding it in both solutions:

	$$\forall_{r,c}: \sum_{\substack{r-1\le r' \le r+1\\c-1\le c'\le c+1}}\text{grid}_{r',c'} = \sum_{\substack{r-1\le r' \le r+1\\c-1\le c'\le c+1}}\text{filled}_{r',c'}$$
	This formula does not handle the edges, but we do handle those cases in our code. Since we are interested in the positions of the numbers in the puzzle, we track these constraints so that Z3 will give them as part of the unsat core.

	Now that we have all constraints we ask Z3 to check if there is a solution.
	If there is a solution we print this alternate solution.
	There exist desired solutions for which this happens, but it is rare.

	Otherwise we ask Z3 for an unsat core. This gives us a list of the number constraints that guarantee there is no second solution. Since these constraints correspond with numbers in the grid, we have found a puzzle that satisfies requirements (1) and (2). We enabled the \texttt{core.minimize} flag of Z3 to try find a minimal puzzle. However the minimal unsat core of Z3 is "best-effort", and there is no guarantee that it is optimal. In particular, the example puzzle above has been generated with this program and has 15 numbers. However, our other program has found the following puzzle with 13 hints:

	\begin{figure}[H]
		\centering
		\includesvg[width=0.8\textwidth]{resources/mosaic-recursive}
	\end{figure}
	% subsection mosaic_unsat_core (end)

	\subsection{Generation using iteration} % (fold)
	\label{sub:mosaic_iteration}
	In this approach we use two Z3 optimizers.
	The first optimizer searches for constraints on where to place numbers and makes these constraints as small as possible.
	The second optimizer tries to find the smallest placement of numbers that satisfies the constraints generated by the first optimizer.\\

	The idea of the first optimizer that we try to find a puzzle that satisfies all the generated constraints but still has a different solution. Then we consider which hints can differentiate both solutions. These are exactly the cells where the count of surrounding black squares differ.
	Any puzzle with the desired solution as it's only solution must then have a number in at least one of those squares.\\

	The constraints we add to the first optimizer largely correspond with what we did for the unsat core. However, we now have three types of variables:

	\begin{tabular}{ll}
		hinted & a puzzle that satisfies the generated constraints\\
		filled & an alternate solution to that puzzle\\
		contributes & places where the count of surrounding black squares differ.
	\end{tabular}

	Again we encode these variables with a 0-1 constrained integer in Z3:

	$$\forall_{r,c}: 0\le \text{hinted}_{r,c} \le 1$$
	$$\forall_{r,c}: 0\le \text{filled}_{r,c} \le 1$$
	$$\forall_{r,c}: 0\le \text{contributes}_{r,c} \le 1$$

	We replace the number constraint with a new one that takes into account whether a number is placed into a cell:

	$$\forall_{r,c}: \text{hinted}_{r,c}\to\sum_{\substack{r-1\le r' \le r+1\\c-1\le c'\le c+1}}\text{grid}_{r',c'} = \sum_{\substack{r-1\le r' \le r+1\\c-1\le c'\le c+1}}\text{filled}_{r',c'}$$

	As before we require that the alternate solution differs in at least one cell from the desired solution:

	$$\exists_{r,c}:  \text{filled}_{r,c} \ne \text{grid}_{r,c}$$

	Next we encode that a cell differentiates both solutions when the black counts differ:

	$$\forall_{r,c}: \sum_{\substack{r-1\le r' \le r+1\\c-1\le c'\le c+1}}\text{grid}_{r',c'} = \sum_{\substack{r-1\le r' \le r+1\\c-1\le c'\le c+1}}\text{filled}_{r',c'}\to \text{contributes}_{r,c}=1$$

	The clause we generate has a literal for every cell that distinguishes both solutions. Since SMT solvers are generally faster on smaller clauses, we minimize the number of cells that distinguish both solutions.

	$$\sum_{r,c}\text{constributes}_{r,c} = \min_{}(\sum_{r,c}\text{contributes}_{r,c}
	)$$

	When we run the first optimizer, we can have two results.
	If the solver returns unsat, there is no puzzle that satisfies all generated constraints and has multiple solutions. In other words, all puzzles satisfying the generated constraints have a unique solution. We then use the second optimizer to find the smallest puzzle satisfying all generated constraints.\\

	In the other case the solver provides a model that includes a puzzle, an alternate solution and the places where the black counts differ. We will add a new clause to both optimizers that encodes that a puzzle should distinguish both solutions:

	$$\exists_{r,c : \text{model}(\text{contributes}_{r,c})=1}: \text{hinted}_{r,c}$$

	After adding this constraint we run the first optimizer again to see if there are still alternate solutions.\\

	The second optimizer is a lot easier.
	It only uses the hinted variables, and the constraint that these are 0 or 1.
	It also includes all generated constraints to guarantee that the resulting puzzle has an unique solution.
	Finally it has the goal to minimize the number of cells that are hinted.\\

	This optimizer will always give a model\footnote{Except in the rare cases where even the puzzle with a number in every cell has multiple solutions}.
	This model corresponds to a minimal puzzle with a unique solution. Suppose there is another puzzle that has fewer hints. Then this puzzle does not satisfy at least one of the generated constraints. But then the alternate solution we found in the step that generated this constraint is also an alternate solution to the puzzle. Therefore there are no smaller puzzles with a unique solution.
	% subsection mosaic_iteration (end)

	\subsection{Results} % (fold)
	\label{sub:results}
	We have discussed two approaches to generating minimal mosaic puzzles for a given solution.
	The unsat-core approach is more efficient since it uses features of Z3 directly. We were able to generate puzzles with a $20\times20$ grid within a couple of seconds. Unfortunately there is no guarantee that the generated puzzle is minimal, as we have shown with the example above.\\

	The iterative approach is quite slow. It took multiple minutes to find the puzzle that disproved minimality of the unsat-core puzzle. The largest puzzle we have been able to generate is $8\times 8$, and took a couple of hours of computation. We are however confident that it generates globally minimal puzzles. An added benefit of this approach is that it defines the space of unique puzzles, so we can easily adapt the goal to find unique puzzles with some other property. For example, we can look for a minimal puzzle that has no adjacent hints, something we cannot encode in the unsat-core approach.
	% subsection results (end)

% section mosaïc (end)

\appendix
\pagebreak
\section{Groups}
\label{apx: groups}
\begin{enumerate}[label=\alph*)]
	\item
		Smallest finite group for which $inv(x*y)=inv(x)*inv(y)$ does not hold.
		\begin{verbatim}
		interpretation( 6, [number=1, seconds=0], [
